---
title: "Causal Pitchfork - The Data"
subtitle: "Data-based storytelling"
author: 
- Daniel Winkler
- Nils Wlömert
format: 
  html:
    code-fold: true
    table-of-contents: true
date: "Updated: `r Sys.Date()`"
bibliography: data_storytelling.bib
---

```{r}
#| include = FALSE
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warnings = FALSE,
  fig.width = 6, fig.height = 6, fig.align = "center"
)
```

# Introduction


This document deals with a fundamental question of causal inference: **Which variables should be included in a causal model?** To answer this question two points need to be clear:

1. In general each causal model only investigates the causal effect of a single independent variable, $x_k$, on the dependent variable $y$. The coefficients associated with all other variables, $x_{j\neq k}$, cannot (automatically) be interpreted as causal relationships. As regression coefficients are commonly presented in a single table, it is often unclear to the reader which coefficients can be interpreted as causal [see @westreich2013table].
2. Statistical significance (or any other statistical test) does not give us any idea about the causal model. To illustrate this, the following figure shows three statistically significant relationships between the variables $x$ and $y$ (all t-stats $> 9$). However, by construction there is no causal relationship between them in two of these examples. Even more concerning: In one case the _exclusion_ of a control variable leads to spurious correlation (leftmost plot) while in the other the _inclusion_ of the control variable does the same (rightmost plot).


```{r intro}
#| fig-width: 12
library(tidyverse)
library(patchwork)
set.seed(11)
## Fork
# n ... number of observations
n <- 500
# d ... binary confounder
d <- rbinom(n, 1, 0.5)
x <- 1.5 * d + rnorm(n)
y <- 0.4 + 2 * d + rnorm(n)
data_fork <- data.frame(x, y, d = factor(d, levels = c(0, 1), labels = c("Yes", "No")))
plt_fork <- ggplot(data_fork, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  ggtitle("Relation due to omitted confounder")
## Pipe
set.seed(11)
x <- 1 * rnorm(n)
z <- rbinom(n, 1, boot::inv.logit(2 * x + rnorm(n)))
y <- 2 * z + rnorm(n)
data_pipe <- data.frame(x, z = factor(z, levels = c(0, 1), labels = c("Yes", "No")), y)
plt_pipe <- ggplot(data_pipe, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  ggtitle("Relation through mediator")
## Collider
set.seed(11)
x <- rnorm(n)
y <- rnorm(n)
a <- rbinom(n, 1, boot::inv.logit(9 * x - 9 * y + rnorm(n)))
data_collider <- data.frame(x, y, a = factor(a, levels = c(0, 1), labels = c("No", "Yes")))
data_collider$x_a <- resid(lm(x ~ 0 + a))
data_collider$y_a <- resid(lm(y ~ 0 + a))
plt_collider <- ggplot(data_collider, aes(x_a, y_a)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(x = "x", y = "y") +
  theme(legend.position = "top") +
  ggtitle("Relation due to included collider")
plt_fork + plt_pipe + plt_collider
```


# The Fork (Good control)

```{r fork}
#| fig.height = 2.5
set.seed(42)
library(ggdag)
library(gt)
library(dagitty)
confounder <- dagify(x ~ d, y ~ d,
  coords = list(
    x = c(x = 1, y = 2, d = 1.5),
    y = c(x = 1, y = 2, d = 2)
  )
) |>
  tidy_dagitty() |>
  mutate(fill = ifelse(name == "d", "Confounder", "variables of interest")) |>
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_point(size = 7, aes(color = fill)) +
  geom_dag_edges(show.legend = FALSE) +
  geom_dag_text() +
  theme_dag() +
  theme(
    legend.title = element_blank(),
    legend.position = "top"
  )
confounder
```

A typical dataset with a confounder will exhibit correlation between the treatment $X$ and outcome $y.$ This relationship is not causal! In the example below we have a binary confounder $d$ (Yes/No) that is d-connected with both $X$ and $y$ ($X$ and $y$ are not d-connected) 

```{r fork_no}
set.seed(11)
# n ... number of observations
n <- 500
# d ... binary confounder
d <- rbinom(n, 1, 0.5)
x <- 1.5 * d + rnorm(n)
y <- 0.4 + 2 * d + rnorm(n)
data_fork <- data.frame(x, y, d = factor(d, levels = c(0, 1), labels = c("Yes", "No")))
ggplot(data_fork, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal()
lm(y ~ x) |>
  broom::tidy() |>
  gt() |>
  fmt_number(
    columns = estimate:p.value,
    decimals = 4
  )
```

However once we take the confounder into account the association vanishes which reflects the lack of a causal relationship in this case (note that for simplicity the regression lines in the plot are not the same as the model output shown). 

```{r fork_yes}
# options(scipen = 10)
ggplot(data_fork, aes(x, y, color = d)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  theme(legend.position = "top")
lm(y ~ x * d, data_fork) |>
  broom::tidy() |>
  gt() |>
  fmt_number(
    columns = estimate:p.value,
    decimals = 4
  )
```

# The Pipe (Bad control)

```{r pipe, fig.height = 2}
med <- dagify(z ~ x, y ~ z,
  coords = list(x = c(x = 1, z = 1.5, y = 2), y = c(x = 1, y = 1, z = 1))
) |>
  tidy_dagitty() |>
  mutate(fill = ifelse(name == "z", "Mediator", "variables of interest")) |>
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_point(size = 7, aes(color = fill)) +
  geom_dag_edges(show.legend = FALSE) +
  geom_dag_text() +
  theme_dag() +
  theme(
    legend.title = element_blank(),
    legend.position = "top"
  )
med
```

If we have a mediator in our data the picture looks very similar to the previous one. In addition, taking the mediator into account also has a similar effect: we remove the association between $X$ and $y$. However, in this case that is not what we want since $X$ and $y$ are d-connected. $X$ causes $y$ through $z$ (note that for simplicity the regression lines in the second plot are not the same as the model output shown).


```{r pipe_no}
set.seed(11)
x <- 1 * rnorm(n)
z <- rbinom(n, 1, boot::inv.logit(2 * x + rnorm(n)))
y <- 2 * z + rnorm(n)
data_pipe <- data.frame(x, z = factor(z, levels = c(0, 1), labels = c("Yes", "No")), y)
ggplot(data_pipe, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal()
lm(y ~ x) |>
  broom::tidy() |>
  gt() |>
  fmt_number(
    columns = estimate:p.value,
    decimals = 4
  )
```


```{r pipe_yes}
ggplot(data_pipe, aes(x, y, color = z)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  theme(legend.position = "top")
lm(y ~ x * z) |>
  broom::tidy() |>
  gt() |>
  fmt_number(
    columns = estimate:p.value,
    decimals = 4
  )
```

# The Collider (Bad control)

```{r}
#| fig.height = 2.5
dagify(a ~ x, a ~ y,
  coords = list(x = c(x = 1, y = 2, a = 1.5), y = c(x = 1, y = 0,  a = 0))
) |>
  tidy_dagitty() |>
  mutate(fill = ifelse(name == "a", "Collider", "variables of interest")) |>
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_point(size = 7, aes(color = fill)) +
  geom_dag_edges(show.legend = FALSE) +
  geom_dag_text() +
  theme_dag() +
  theme(
    legend.title = element_blank(),
    legend.position = "top"
  )
```

The collider is a special case. There is no association between $X$ and $y$ as long as we do **not** account for the collider in the model. However, by accounting for the collider we implicitly learn about $y$ as well (we use $X$ as the predictor). Since the collider is caused by $X$ and $y$, we can figure out what $y$ must be once we know $X$ and the collider similar to solving a simple equation you would see in high-school.

```{r}
set.seed(11)
x <- rnorm(n)
y <- rnorm(n)
a <- rbinom(n, 1, boot::inv.logit(9 * x - 9 * y + rnorm(n)))
data_collider <- data.frame(x, y, a = factor(a, levels = c(0, 1), labels = c("No", "Yes")))
ggplot(data_collider, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal()
lm(y ~ x) |>
  broom::tidy() |>
  gt() |>
  fmt_number(columns = estimate:p.value, decimals = 4)
```

```{r}
data_collider$x_a <- resid(lm(x ~ 0 + a))
data_collider$y_a <- resid(lm(y ~ 0 + a))
ggplot(data_collider, aes(x_a, y_a)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(x = "x after accounting for a", y = "y after accounting for a") +
  theme(legend.position = "top")
lm(y ~ x + a, data_collider) |>
  broom::tidy() |>
  gt() |>
  fmt_number(columns = estimate:p.value, decimals = 4)
```

# Connections to related concepts

## Omitted Variable Bias (OVB)

[Recall](https://wu-rds.github.io/MA2022/regression.html#omitted-variables) that variables that influence both the outcome and other independent variables will bias the coefficients of those other independent variables if left out of a model. This bias is referred to as "Omitted Variable Bias" (short OVB) since it occurs due to the omission of a crucial variable. OVB occurs whenever a confounder (see [The Fork](#the-fork-good-control)) is left out of the model. 

## Mediation analysis

Notable changes to [The Pipe](#the-pipe-bad-control): 

- We have both direct and indirect effects of $x$ on $y$
- The mediator $m$ is continuous instead of binary


```{r}
#| fig-height: 2.5
med2 <- dagify(m ~ x, y ~ m + x,
  coords = list(x = c(x = 1, m = 1.5, y = 2), y = c(x = 1, y = 1, m = 1.5))
) |>
  tidy_dagitty() |>
  mutate(fill = ifelse(name == "m", "Mediator", "variables of interest")) |>
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_point(size = 7, aes(color = fill)) +
  geom_dag_edges(show.legend = FALSE) +
  geom_dag_text() +
  theme_dag() +
  theme(
    legend.title = element_blank(),
    legend.position = "top"
  )
med2
```


```{r}
#| code-fold: false
set.seed(11)
X <- 100 * rnorm(n)
M <- 10 + 0.5 * X + 5 * rnorm(n)
Y <- -25 + 4 * X + 3 * M + rnorm(n)
total_effect <- lm(Y ~ X)
summary(total_effect)
avg_direct_effect <- lm(Y ~ X + M)
summary(avg_direct_effect)
X_on_M <- lm(M ~ X)
summary(X_on_M)
avg_causal_mediation_effect <- coef(X_on_M)['X'] * coef(avg_direct_effect)['M']
cat("Average Causal Mediation Effect (ACME):", avg_causal_mediation_effect)
cat("Average Direct Effect (ADE):", coef(avg_direct_effect)['X'])
cat("Total Effect:", coef(total_effect)['X'])
cat("Also Total Effect:", coef(avg_direct_effect)['X'] + avg_causal_mediation_effect)
cat("Proportion Mediated:", avg_causal_mediation_effect / coef(total_effect)['X'])
```


```{r}
#| message: false
library(mediation)
mediation_result <- mediate(X_on_M, avg_direct_effect, 
                            treat = 'X', mediator = 'M',
                            boot = TRUE, sims = 1000)
summary(mediation_result)
```

Equivalent to PROCESS model 4 (requires sourcing the `process.R` file that can be downloaded [here](https://haskayne.ucalgary.ca/CCRAM/resource-hub)):

```{r}
#| eval: false
#| code-fold: false
process(data.frame(Y, X, M), y = 'Y', x = 'X', m = 'M', model = 4)
```

Alternatively the following code _should_ download and source the macro for you but will definitely break in the future (try changing the `v42` part of the link to `v43` or `v44` etc. or obtain a new link from [the website](https://haskayne.ucalgary.ca/CCRAM/resource-hub) if it does):

```{r}
#| eval: false
#| code-fold: false
temp <- tempfile()
download.file("https://www.afhayes.com/public/processv42.zip",temp)
files <- unzip(temp, list = TRUE)
fname <- files$Name[endsWith(files$Name, "process.R")]
source(unz(temp, fname))
unlink(temp)
```



# Appendix: How the sausage is made

The fork, mediator, and collider were generated as binary variables to make visualization easier. Binary variables can be drawn from a so-called Bernoulli distribution which is a special case of the binomial distribution with size = 1.
The distribution takes the probability of getting a $1$ as input.

## The Fork

```{r}
#| code-fold: false
## Make code reproducible
set.seed(11)
## Number of observations
n <- 1500
## Random draw from Bernoulli with p(1) = 0.5, p(0) = 0.5
d <- rbinom(n, 1, 0.5)
## X is caused by d
x <- 2 * d + rnorm(n)
## y is caused by d
y <- 0.4 + 2.5 * d + rnorm(n)
fork <- data.frame(x, y, d = factor(d,
  levels = c(0, 1),
  labels = c("No", "Yes")
))
ggplot(fork, aes(x, y, color = d)) +
  geom_point() +
  theme_minimal() +
  theme(legend.position = "top")
```

## The Pipe

```{r, echo = TRUE, fig.height = 4}
## Generate random X
x <- rnorm(n)
## inv.logit ensures that values are between 0 and 1
ggplot(data.frame()) +
  stat_function(fun = boot::inv.logit, xlim = c(-10, 10)) +
  theme_minimal() +
  labs(title = "Inverse Logit function", x = "x", y = "inv.logit(x)")
```

```{r, echo = TRUE}
## z is caused by X
z <- rbinom(n, 1, boot::inv.logit(2 * x + rnorm(n)))
## y is caused by z
y <- 2 * z + rnorm(n)
pipe <- data.frame(x, y, z = factor(z,
  levels = c(0, 1),
  labels = c("Yes", "No")
))
ggplot(pipe, aes(x, y, color = z)) +
  geom_point() +
  theme_minimal() +
  theme(legend.position = "top")
```

## The Collider

```{r, echo = TRUE}
## Generate random x
x <- rnorm(n)
## Generate random y
y <- rnorm(n)
## a is caused by both X and y
a <- rbinom(n, 1, boot::inv.logit(9 * x - 9 * y + rnorm(n)))
collider <- data.frame(x, y, a = factor(a,
  levels = c(0, 1),
  labels = c("No", "Yes")
))
ggplot(collider, aes(x, y)) +
  geom_point() +
  theme_minimal()
```

In order to get the partial correlation of $X$ and $y$ after accounting for $a$ we first regress both $X$ and $y$ on $a$ and use the unexplained part (residual) in the plot. This is equivalent to a regression that has both $X$ and $a$ as explanatory variables.

```{r, echo = TRUE, fig.width=7, fig.height=7}
collider$x_a <- residuals(lm(x ~ 0 + a))
collider$y_a <- residuals(lm(y ~ 0 + a))
ggplot(collider, aes(x_a, y_a)) +
  geom_point() +
  theme_minimal() +
  labs(x = "x after accounting for a", y = "y after accounting for a")
```
